Sophize Mathematics Library

### 2022-05-27 math_library/sophize_math.tex
Latex package for CICM '22 submission

\documentclass[a4paper]{article}
\usepackage{graphicx}
\usepackage{onecolceurws}
\usepackage{csquotes}
\usepackage{hyperref}
\usepackage{framed}
\usepackage{mdframed}
\usepackage{upquote}
\usepackage{lmodern}
\usepackage[normalem]{ulem}

\newcommand\todo[1]{\textcolor{red}{TODO: #1}}

\title{Sophize Knowledge Organization}

\author{ Abhishek Chugh }

\graphicspath {{../res}}

\institution{ Sophize Foundation\\ Bengaluru, India \\ abc@sophize.org }

\begin{document}
\maketitle

\begin{abstract}
Sophize is a novel mathematics library and discussion platform with a mission to help our users find and organize mathematical proofs. We present Sophize's novel yet familiar knowledge organization scheme that is used to represent a wide variety proofs. With this knowledge organization scheme at its core, we have engineered a platform to aggregate knowledge from multiple sources. The platform meaningfully combines knowledge from published research, encyclopaedias such as PlanetMath and Wikipedia, informal computer programs and formal systems such as Metamath; and allows its users to run a federated search over them. To make this happen, two developments were necessary. First, we developed an open plugin architecture that allows proofs from computer programs to be generated as required. Second, we extended the Markdown language to represent the connections between between mathematical objects found across various sources of knowledge. In addition, we also utilized this new language to create a novel communication system built specifically to aid mathematicians in solving problems collaboratively.

\end{abstract}

\vskip 32pt

@@

\subsubsection*{Background: The Sophize Platform}

- The goal is rational truth. Three dimensions are consistency, correspondence and values.

- No hidden assumptions. Should have clear answer to every question of the form `why do you believe X'.

- In its current iteration, Sophize implements consistency. Knowledge represented by manually and automated generation of arguments. Can model vast majority (almost all) of mathematics.

- This gives you a way to create (consistency enforced) models. Next step is to see how accurately these models represent (parts of) the World. That involves quantifying correspondence of observations with the predictions made by the theories.

- And finally, you need model individual, institutional, and societal values to claim the truth of `ought statements' like `I should wear a mask' or `Masks should be mandated'.

- These starting beliefs (axioms, principles, laws, assumptions) and values are a way to model our belief systems. There are no absolutes, each belief system results in a different rational truth.

\subsubsection*{Proofs in Sophize}
Sophize is a novel mathematics library and discussion platform. The author has developed the platform over the course of the last two years at \url{https://sophize.org}. Sophize's primary mission is to help users find existing proofs of mathematical statements, to discover new proofs, and to utilize this knowledge in their work. We combine knowledge from multiple resources and have accumulated thousands of definitions, theorems, and proofs from a wide variety of sources, including some published research, the PlanetMath encyclopaedia, Wikipedia, and the Metamath formal system \cite{metamath}.

@@

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=12cm]{proof_tree}}
\caption{A proof displayed as a graph (tree) of arguments and propositions on the Sophize platform.}
\label{proof_tree}
\end{center}
\end{figure}

@@

Hence, we required novel knowledge organization techniques to connect mathematical entities from such a wide range of formal and informal knowledge sources. And thus, one of the main challenges towards building the Sophize platform was to create a simple way to embed mathematical entities on the web. \todo{Explain and summarize the problem statement}

\subsubsection*{Contribution}
\todo{summarize sections. section 3 does X, section 4 does Y etc.}

@@

We use the set of all known arguments to track knowledge from a variety of foundational theories. For theoretical knowledge, i.e., knowledge that doesn't involve empirical data, the primary criteria of validity is internal (logical) consistency. Thus, for each foundational theory (`belief set' described below), we also keep track of any contradictions that arise from those foundations.
\todo{section is crap, overview of how aggregated}

@@

\subsection{Argument Graph}
An argument can be seen as a simple graph (see Figure \ref{argument}) with two types of nodes - (a) a single node representing the argument itself and (b) a set of proposition nodes representing the argument's conclusion and premises. The edges of this graph are directed and go from (a) premise nodes to argument node and (b) argument node to conclusion node.

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=3.5cm]{argument}}
\caption{An argument represented as a graph.}
\label{argument}
\end{center}
\end{figure}

A set of arguments can thus be represented as a single graph that we call an \textbf{argument graph} (Figure \ref{argument_graph}).

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=7cm]{argument_graph}}
\caption{A set of arguments forming an argument graph.}
\label{argument_graph}
\end{center}
\end{figure}

\subsection{Generating Proof Graphs}
An argument graph along with a set of set of axioms results in a \textbf{proof graph}. Any reachability algorithm starting with the axiom nodes is sufficient to create the proof graph (See Figure \ref{proof_graph1}).

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=7cm]{proof_graph1}}
\caption{Proof graph from argument graph (Figure \ref{argument_graph}) using `Prop 2', `Prop 3', Prop 4' are axioms.}
\label{proof_graph1}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=6cm]{proof_prop7}}
\caption{Proof of `Prop 7' from proof graph in Figure \ref{proof_graph1} uses arguments `Arg 4' and `Arg 5'.}
\label{proof_prop7}
\end{center}
\end{figure}

Note that an argument graph combines with different sets of axioms to generate different proof graphs. The same argument may be utilized in multiple belief sets (Figure \ref{proof_graph2}).

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=8cm]{proof_graph2}}
\caption{Proof graph from the same argument graph (Figure \ref{argument_graph}) using `Prop 1', `Prop 2', and Prop 7' as axioms.}
\label{proof_graph2}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=5cm]{validity_criterion}}
\caption{Validity criteria as an argument premise.}
\label{validity_criterion}
\end{center}
\end{figure}

@@

\subsection{Aggregating Knowledge}

@@

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=5cm]{modal}}
\caption{Modal dialog box for the term binary operation. These boxes can also have content that uses all three types of links. }
\label{modal}
\end{center}
\end{figure}

@@

Sophize markdown adds an icon next to each proposition link. This icon indicates whether the proposition has been proven or dis-proven (or both) in the currently selected belief set using a check-mark or a cross. Clicking on the icon brings up the proof graph as shown in Figure \ref{proof_tree}. If the users switches from one belief set to another, the icon is updated based on the truth values in the new belief set but other parts of the article are unchanged.

@@

\begin{figure}[htbp]
\begin{center}
\fbox{\includegraphics[height=3cm]{formal}}
\caption{The prime number theorem written in Metamath language and rendered as Markdown in real time.}
\label{formal}
\end{center}
\end{figure}

\section{Building a Modern Mathematics Library}
The argument and proof graphs in help us build a network of information that spans across a wide variety of sources (including live results from PGMs). Using different belief sets, users can view mathematical knowledge arising from different theoretical foundations and choose what they consider to be reliable. With Sophize Markdown, users can easily embed semantic information, thus making it convenient to explore the network of knowledge that is being built. However, a few more practical issues need to be overcome to build a modern scalable library with reliable information.

@@

\subsubsection{}

\subsubsection{Connecting Across Datasets}
The sub-problem 4 mentioned above is crucial to help make connections across datasets. For example, once we have identified that the `Jensen's Inequaility' in a published paper is same as the Planetmath proposition with the same name, Planetmath's proof is made available to the reader of the published paper.

\subsection{Search Interface}

\subsection{Data Management}
Sophize organizes

@@

\todo{copyright/access management}

\todo{efficient computation and storage of proof graphs}

\todo{ storing and reading data in disk/GitHub like repo }

With Sophize Markdown at its core, Sophize's collaboration interface\cite{todo}  (\underline{\href{https://youtu.be/d3gaalJ7UQM}{demo}}) is able to provide key features such as easy linking of existing math content, \LaTeX\space support, and live preview comment drafts.

\subsection{Connection across data sources}
\todo{What are Translations? (informal to formal, formal to other formal), Why do we need translations?}

There is no good way to claim equivalence of propositions across different formal systems. [Formal Harmony still in infancy, theory morphisms aren't yet practical].

Eg. ZFC (informal) + ZFC (metamath) + Jensen's Inequality (metamath) + validity criteria (translation feels right to metamath group) $\rightarrow$ 
Jensen's Inequality (informal)

But, we can represent translations manually using arguments. Such translations do not stand on a very firm foundation and that should be indicated in the validation criteria of the argument. Again, a user can choose not to accept such criteria. In that case, these translations will not be used in the generating proof graphs.

Thus we can create proof graphs that have arguments from different languages.

\subsection{Uses}
Efficiently modelling different axiomatic systems that have a large overlap of arguments(eg. euclidean/parabolic/hyperbolic geometry, ZF/ZFC etc)

\subsubsection{Other Notes}
\todo{What criteria are met based on the 2013 report}

@@

\bibliographystyle{alpha}

\bibliography{mybib}

\end{document}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
\usepackage{mdframed}
\usepackage[normalem]{ulem}
\usepackage{hyperref}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}
%

\newcommand\todo[1]{\textcolor{red}{TODO: #1}}
\graphicspath {{../res}}
\begin{document}
%
\title{Sophize Mathematics Library}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Abhishek Chugh\inst{1}\orcidID{0000-0001-6765-5229}}
%
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Sophize Foundation, Bengaluru - 560066, India 
\email{abc@sophize.org}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Sophize is a novel mathematics library and discussion platform with a mission to help our
users find and organize mathematical proofs. We present Sophize's novel yet familiar
knowledge organization scheme that is used to represent a wide variety of proofs. With this
scheme at its core, we have engineered a platform to logically aggregate knowledge from
multiple sources. The platform curates knowledge from published research, encyclopaedias
such as PlanetMath and Wikipedia, informal computer programs, and formal systems such as
Metamath; and allows its users to run a federated search over them. To make this happen, two
developments were necessary. First, we came up with the concept of 'proof-generating
machines' and developed an open plugin architecture that allows proofs from computer
programs to be generated as required. Second, we extended the Markdown language to represent
the connections between mathematical objects found across various sources of knowledge. In
addition, we also utilized this new language to create a novel communication system built
specifically to aid mathematicians in solving problems collaboratively.

\keywords{Sophize \and
knowledge organization \and
proof-generating machine \and
belief set \and
metamath \and
markdown \and
argument graph \and
proof graph \and
semantic data}
\end{abstract}

@@

\subsection*{The Sophize Platform}

Sophize is a novel mathematics library and discussion platform. The author has developed the platform over the course of the last two years at \url{https://sophize.org}. Sophize's primary mission is to help users find existing proofs of mathematical statements, to discover new proofs, and to utilize this knowledge in their work. We combine knowledge from multiple resources and have accumulated thousands of definitions, theorems, and proofs from a wide variety of sources, including some published research, the PlanetMath encyclopedia, Wikipedia, and the Metamath formal system \cite{metamath}.

\subsection*{Motivation}

In the past couple of decades, the Internet has revolutionized our life by giving us access to seemingly unlimited information. Information from a wide variety of topics ranging from human medicine to rocket science is now just a few keystrokes away. However, even with this information we seem to having trouble understanding the complex world around us. In fact, dealing with the vast amounts of information thrown at us seems to be one of the primary problems of our generation. While we have access to a wide range of opinions on almost every topic imaginable, it has becoming increasingly difficult to decide which one to trust. It seems that you can find "proofs" for anything and everything we would like to believe.

Thus the author's primary motivation for this project is to create a library that not just lays out all the information of the World but organizes in a way that help us make better sense of it. A library that helps us access justifications from a variety of viewpoints and help us collaboratively and objectively evaluate issues in the justifications or the viewpoints they arise from. Clearly, this is challenging goal.

We begin our journey with a focus on pure Mathematics, where knowledge is objectively less complex as compared to other fields. In mathematics, only logically sound arguments are acceptable as justifications. These arguments develop from different foundational theories and logical consistency is the primary criteria of correctness for them. Thus, as a Mathematics library, Sophize's goal is to help our users access to mathematical proofs (chains of arguments) from a variety of foundational theories (extended to `belief sets' as described in Section \ref{sec:bset}) and help us collaboratively and objectively evaluate issues in the argument or the belief sets they arise from.

With our current scheme we can incorporate almost all mathematics. A natural extension would be model observational data and develop criteria to quantify empirical correspondence in order to have the ability to incorporate scientific knowledge. But, for now we are focused on building a state-of-the-art Mathematics library that is valuable to our community.

\subsection*{Proofs in Sophize}

\begin{figure}[!ht]
\begin{center}
\fbox{\includegraphics[width=\textwidth]{proof_tree}}
\caption{A proof displayed as a graph (tree) of arguments and propositions on the Sophize platform.}
\label{proof_tree}
\end{center}
\end{figure}

@@

@@

Hence, we required novel knowledge organization techniques to connect mathematical entities from such a wide range of formal and informal knowledge sources. And thus, one of the main challenges towards building the Sophize platform was to create a simple way to embed mathematical entities on the web.

\subsection*{Contribution}
In this paper, we present the novel concepts and features of Sophize Mathematics Library. First, we expound Sophize's scheme to build a trust-focused logical network of mathematical knowledge. In section \ref{sec:pgm}, we describe Sophize's proof-generating machines which are computer programs that generate proofs that integrate with Sophize's knowledge graphs.

We then present some features of Sophize Markdown, an extension of the Markdown language that helps present Sophize's knowledge graphs to its users. It is convenient enough to be used for casual discussions of mathematical ideas over the web. It is also powerful enough to embed mathematical entities such as definitions, theorems, and proofs from various sources, including formal systems. It thus plays an integral role in the two problems that we have mentioned above.

Section \ref{sec:lib} discusses some other challenges and contributions made towards making Sophize a modern mathematics library. Finally, Section \ref{sec:summary} concludes the paper.

@@

We use the set of all known arguments to track knowledge from a variety of foundational theories. For theoretical knowledge, i.e., knowledge that doesn't involve empirical data, the primary criteria of validity is internal (logical) consistency. Thus, for each foundational theory (`belief set' described below), we also keep track of any contradictions that arise from those foundations.

@@

\subsection{Argument Graph}
An argument can be seen as a simple graph (see Fig.~\ref{argument}) with two types of nodes - (a) a single node representing the argument itself and (b) a set of proposition nodes representing the argument's conclusion and premises. The edges of this graph are directed and go from (a) premise nodes to argument node and (b) argument node to conclusion node.

\begin{figure}[!ht]
\begin{center}
\fbox{\includegraphics[height=3cm]{argument}}
\caption{An argument represented as a graph.}
\label{argument}
\end{center}
\end{figure}

A set of arguments can thus be represented as a single graph that we call an \textbf{argument graph} (Fig.~\ref{argument_graph}).

\begin{figure}[!ht]
\begin{center}
\fbox{\includegraphics[height=5.5cm]{argument_graph}}
\caption{A set of arguments forming an argument graph.}
\label{argument_graph}
\end{center}
\end{figure}

\subsection{Generating Proof Graphs}
An argument graph along with a set of set of axioms results in a \textbf{proof graph}. Any reachability algorithm starting with the axiom nodes is sufficient to create the proof graph (See Fig.~\ref{proof_graph1}).

\begin{figure}[!ht] 
\begin{center}
\fbox{\includegraphics[height=5.5cm]{proof_graph1}}
\caption{Proof graph from argument graph (Fig.~\ref{argument_graph}) using `Prop 2', `Prop 3', Prop 4' are axioms.}
\label{proof_graph1}
\end{center}
\end{figure}

@@

\begin{figure}[!ht]
\begin{center}
\fbox{\includegraphics[height=4cm]{proof_prop7}}
\caption{Proof of `Prop 7' from proof graph in Fig.~\ref{proof_graph1} uses arguments `Arg 4' and `Arg 5'.}
\label{proof_prop7}
\end{center}
\end{figure}

Note that an argument graph combines with different sets of axioms to generate different proof graphs. The same argument may be utilized in multiple belief sets (Fig.~\ref{proof_graph2}).

\begin{figure}[!ht]
\begin{center}
\fbox{\includegraphics[height=5.5cm]{proof_graph2}}
\caption{Proof graph from the same argument graph (Fig.~\ref{argument_graph}) using `Prop 1', `Prop 2', and Prop 7' as axioms.}
\label{proof_graph2}
\end{center}
\end{figure}

@@

\begin{figure}[!ht]
\begin{center}
\fbox{\includegraphics[height=4cm]{validity_criterion}}
\caption{Validity criteria as an argument premise.}
\label{validity_criterion}
\end{center}
\end{figure}

@@

\subsection{Advantages and Limitations}

Efficiently modelling different axiomatic systems that have a large overlap of arguments(eg. euclidean/parabolic/hyperbolic geometry, ZF/ZFC etc)

@@

\begin{figure}
\begin{center}
\fbox{\includegraphics[width=\textwidth]{modal}}
\caption{Modal dialog box for the term binary operation. These boxes can also have content that uses all three types of links. }
\label{modal}
\end{center}
\end{figure}

@@

Sophize markdown adds an icon next to each proposition link. This icon indicates whether the proposition has been proven or dis-proven (or both) in the currently selected belief set using a check-mark or a cross. Clicking on the icon brings up the proof graph as shown in Fig.~\ref{proof_tree}. If the users switches from one belief set to another, the icon is updated based on the truth values in the new belief set but other parts of the article are unchanged.

@@

\begin{figure}
\begin{center}
\fbox{\includegraphics[height=2.25cm]{formal}}
\caption{The prime number theorem written in Metamath language and rendered as Markdown in real time.}
\label{formal}
\end{center}
\end{figure}

\section{Building a Modern Mathematics Library}
\label{sec:lib}
The argument and proof graphs in help us build a network of information that spans across a wide variety of sources (including live results from PGMs). Using different belief sets, users can view mathematical knowledge arising from different theoretical foundations and choose what they consider to be reliable. With Sophize Markdown, users can easily embed semantic information, thus making it convenient to explore the network of knowledge that is being built. However, a few more practical issues need to be overcome to build a modern scalable library with reliable information.

@@

\subsubsection{Connecting Across Datasets}
The last sub-problem mentioned above is crucial to help make connections across datasets. For example, once we have identified that the `Jensen's Inequality' in a published paper is same as the Planetmath proposition with the same name, Planetmath's proof is automatically made available to the reader of the published paper.

However, figuring out if two sources are referring to the same proposition or a different version of the same proposition is often unclear. When working with formal systems using different languages, another related problem arises. There is no general way to claim equivalence of propositions across different formal systems. There are some projects, like Formal Harmony and theory morphisms, that are trying to understand in what sense these propositions are same and how we can translate one to another. But these projects are still in early stages and do not provide a general solution.

However, we can utilize curated translations, such as those available from Logipedia, by representing them using arguments. The conclusion of such an argument is the proposition in target language and the premise would be the proposition in source language. For bi-directional translations two such arguments will be required. Such translations do not stand on a very firm foundation and that should be indicated in the validation criteria of the argument. A user can choose not to accept such criteria. In that case, these translations will not be used in the generating proof graphs. Thus we can create proof graphs that have arguments from different languages.

There are some cases where translations occur on firmer grounds, e.g., use of theory morphisms or translations of HOL-light proofs to Metamath. The author has not studied these cases in detail but expects that it to be possible to formulate such cross-language translations using the argument graphs.

\subsection{Data Management}

Sophize divides its data into groups called datasets for copyright management and data access rules. For example, the 'wiki' dataset that extracts knowledge from Wikipedia mathematics glossaries releases its data under `CC-BY-SA' license, whereas the `metamath' dataset is public (`CC0'). Our signed-in users are free to create their own datasets and we allow them to choose who can read, write or comment in these datasets. Another interesting access control is the restriction of use of resources (especially argument-validity criteria) outside an otherwise open-to-read dataset.

Information can be added to Sophize platform using intuitive web forms. However, it is infeasible to extract information from large repositories in this manner. Sophize publishes its data schema and helper libraries (explained in Section \ref{sec:core}) in multiple languages. The extracted information is can thus be stored as text (JSON) files in a repository. Sophize provides dataset owners the ability to sync knowledge such repositories to the central Sophize database.

\subsection{Search Interface}
Sophize allows its users to run a run a federated search over its resources using an Elasticsearch server. We allow users to search a space-insensitive and case-insensitive search over \LaTeX\space formulae by encoding \LaTeX\space formulae as lexemes and an Elasticsearch analyzer (See section \ref{sources} for sources) that removes spaces and lower cases the encoding.

For example the, latex formula \verb+$E = mc^2$+ is converted to \textit{Ej9rjmc9v2mMm}. 
Note that the encoder converts spaces and new-lines into the character `j' or `J' which are removed by our Elasticsearch analyzer at the time of searching the documents.

@@

The Sophize platform includes an innovative mathematics communication interface designed to help
mathematicians discuss mathematics and collaboratively discover new proofs. The design focuses on
making the collaboration more productive and enjoyable by building the right set of technical
tools that aid in effective organization and summarization of existing progress.
With Sophize Markdown at its core, Sophize's collaboration interface\cite{todo}  (\underline{\href{https://youtu.be/d3gaalJ7UQM}{demo}}) is able to provide key features such as easy linking of existing math content, \LaTeX\space support, and live preview comment drafts.

\section{Conclusion}
\label{sec:conclusion}
\todo{What criteria are met based on the 2013 report}

@@

\subsubsection{Acknowledgements} 
A part of the research presented here was supported by the International Mathematical Union.

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{mybib}
\end{document}
